---
title: "Lake Turkana Isotope Hydrology: New measurements, evaporation model, and mass balance"
date: 2022-03-05
output: 
  html_notebook: 
    highlight: monochrome
  html_document: 
    highlight: monochrome
editor_options: 
  chunk_output_type: inline
reference-section-title: "References"
---

```{r, setup, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(here, knitr, tidyverse, readxl, lubridate, ggthemes, ggrepel, patchwork, kableExtra)
```

```{r plot-styles, echo = FALSE}
# Plot styles 
# Colorblind-friendly palette from @tol2018.

theme_set(theme_tufte(base_family = "Helvetica") +
            theme(
              panel.background = element_rect(fill = 'white'), 
              panel.border = element_rect(fill=NA, color = "black", size = 0.5),
              panel.grid = element_line(color = "gray90", size = 0.5),
              plot.background = element_rect(fill='transparent', color = NA),
              title = element_text(size = 14),
              axis.title = element_text(size = 14),
              axis.text = element_text(size = 14),
              legend.title = element_text(size = 14, face = "bold"),
              legend.text = element_text(size = 14),
              legend.background = element_rect(fill = 'transparent', color = NA), 
              legend.box.background = element_rect(fill = 'transparent', color = NA), 
              legend.position = "right"
            ))

# *To do*: Revisit appropriate axis limits for final plots
# axis.x.max <- max(waters$d18O) + 0.5
# axis.x.min <- min(waters$d18O) - 0.5
# axis.y.max <- max(waters$dD) + 0.5
# axis.y.min <- min(waters$dD) - 0.5

axis.x.label <- expression(paste(delta^{18}, "O (‰ VSMOW)"))
axis.y.label <- expression(paste(delta, "D (‰ VSMOW)"))

color.lake <- "#228833"
color.precip <- "#66CCEE"
color.river <- "#4477AA"
color.ground <- "#EE6677"
color.surface <- "#CCBB44"
color.GMWL <- "gray19"
color.model <- "gray69"

color.scenario1 <- "#DDCC77"
color.scenario2 <- "#CC6677"
color.scenario3 <- "#882255"
color.scenario4 <- "#AA4499"

shape.lake <- 16 # circle
shape.precip <- 17 # triangle
shape.river <- 15 # square
shape.ground <- 7 # x in box
shape.surface <- 5 # diamond empty
shape.surface.lake <- 1 # circle empty
shape.model <- 3 # cross 

size.st <- 3
size.sm <- 1
size.lg <- 5
```

# Water isotope measurements

```{r data-processing, echo = FALSE, message = FALSE}
# Load and process data (this study)
waters.rawdata <- read_csv(here("Turkana-waters/data/2016-2021_waters.csv"), show_col_types = FALSE)
waters <- waters.rawdata |>
  # Simplify WaterType categories
  mutate(WaterType = str_replace_all(WaterType, 
                                     c("River Delta" = "River",
                                       "Deep Ground" = "Ground",
                                       "Shallow Ground" = "Ground",
                                       "Spring" = "Ground",
                                       "Evaporated Surface" = "Surface"))) 

# Load and process data (Levin et al 2009)
waters.Levin.rawdata <- read_excel(here("Turkana-waters/data/Levin_et_al_2009_raw_data.xls"))
waters.Levin <- waters.Levin.rawdata |>
  # Make WaterType categories consistent with this study
  mutate(WaterType = str_replace_all(Type, 
                                     c("spring" = "Ground",
                                       "bore hole" = "Ground",
                                       "river" = "River",
                                       "water hole" = "Ground",
                                       "tap" = "Ground",
                                       "stream" = "Surface",
                                       "well" = "Ground",
                                       "rain" = "Precipitation"))) |>
  # Filter arid latitudes 
  filter(between(Lat, 2, 5.1))

```

## Figure 1. Sample location map

```{r figure1, echo = FALSE, message = FALSE, warning = FALSE}
# Load maps 
source(here("Turkana-maps/scripts/Turkana-base-map.R"))

# Add data
fig1A <- map.Turkana + 
  geom_point(data = waters,
             aes(x = Longitude, y = Latitude, color = WaterType, shape = WaterType), size = size.st) +
  
  # Legend
  scale_shape_manual(
    breaks = c("Lake", "Restricted Lake", "River", "Precipitation", "Surface", "Ground"), 
    # Set plot element shapes
    values = c(
      "Lake" = shape.lake,
      "Restricted Lake" = shape.surface.lake,
      "River" = shape.river,
      "Precipitation" = shape.precip,
      "Surface" = shape.surface,
      "Ground" = shape.ground)) +
  
  scale_color_manual(
    breaks = c("Lake", "Restricted Lake", "River", "Precipitation", "Surface", "Ground"), 
    # Set plot element colors
    values = c(
      "Lake" = color.lake,
      "Restricted Lake" = color.lake,
      "River" = color.river,
      "Precipitation" = color.precip,
      "Surface" = color.surface,
      "Ground" = color.ground))

fig1A + map.context +
  plot_layout(ncol = 2, widths = c(2,1))

# save figures for formatting

```


## Table 1. New measurements

```{r table1, echo = FALSE, message = FALSE, warning = FALSE}
tbl1 <- waters |>
  dplyr::select(SampleID, Date, Latitude, Longitude, Location, WaterType, d18O, dD, Dexcess) 
tbl1

# save Table 1 for Word
#write.csv(tbl1, here("Turkana-waters/figures-tables/Table1.csv"))
```

### Summary values

Data from Levin et al filtered to latitudes between 2 — 5.1°

```{r summary-data, echo = FALSE}
# Filter meteoric waters (all types except ground)
# This study only
waters.meteoric <- waters |>
  filter(WaterType != "Ground")

waters.summary <- waters |>
  group_by(WaterType) |>
  summarize(d18O_mean = mean(d18O),
            d18O_stdv = sd(d18O),
            dD_mean = mean(dD),
            dD_stdv = sd(dD)) |>
  mutate(Source = "this study")

waters.summary.Levin <- waters.Levin |>
  group_by(WaterType) |>
  summarize(d18O_mean = mean(d18O_VSMOW, na.rm = TRUE),
            d18O_stdv = sd(d18O_VSMOW, na.rm = TRUE),
            dD_mean = mean(dD_VSMOW, na.rm = TRUE),
            dD_stdv = sd(dD_VSMOW, na.rm = TRUE)) |>
  mutate(Source = "Levin et al 2009")

Omo.River <- waters.Levin |>
  filter(grepl("Omo River", Sampling_Location_and_Notes)) |>
  mutate(Source = "Levin et al 2009")

mean.surface.TC <- mean(waters$SurfaceTempC, na.rm = TRUE)
mean.rh <- 0.33 # From TBI-Ileret PurpleAir sensor lifetime average

waters.summary <- rbind(waters.summary, waters.summary.Levin) |>
  mutate(d18O_mean = round(d18O_mean, 2),
         d18O_stdv = round(d18O_stdv, 2),
         dD_mean = round(dD_mean, 2),
         dD_stdv = round(dD_stdv, 2)) 

waters.summary |>
  arrange(desc(Source)) 

```

## Figure 2. Water isotope plot 

```{r figure2, echo = FALSE, warning = FALSE}
fig2 <- ggplot() +
  
  geom_point(data = waters.Levin,
             aes(x = d18O_VSMOW, y = dD_VSMOW, shape = WaterType), size = size.st, color = "gray77", 
             show.legend = FALSE) +
  geom_point(data = waters,
             aes(x = d18O, y = dD, color = WaterType, shape = WaterType), size = size.st) +
  geom_abline(data = NULL, aes(slope = 8, intercept = 10, color = "GMWL")) +
  
  labs(x = axis.x.label, 
       y = axis.y.label,
       caption = "Data from Levin et al 2009 (arid latitudes) shown in gray") +
  
  # Legend
  scale_shape_manual(
    breaks = c("Lake", "Restricted Lake", "River", "Precipitation", "Surface", "Ground", "GMWL"), 
    # Set plot element shapes
    values = c(
      "Lake" = shape.lake,
      "Restricted Lake" = shape.surface.lake,
      "River" = shape.river,
      "Precipitation" = shape.precip,
      "Surface" = shape.surface,
      "Ground" = shape.ground,
      "GMWL" = 1)) +
  
  scale_color_manual(
    breaks = c("Lake", "Restricted Lake", "River", "Precipitation", "Surface", "Ground", "GMWL"), 
    # Set plot element colors
    values = c(
      "Lake" = color.lake,
      "Restricted Lake" = color.lake,
      "River" = color.river,
      "Precipitation" = color.precip,
      "Surface" = color.surface,
      "Ground" = color.ground,
      "GMWL" = color.GMWL)) +
  
  guides(
    color = guide_legend(
      override.aes=list(
        linetype = c(rep("blank", 6), "solid"),
        shape = c(shape.lake, shape.surface.lake, shape.river, shape.precip, shape.surface, shape.ground, NA),
        size = 2))) 

plot(fig2)

# save figure for formatting

```


## Figure 3. Time series plots

```{r fig2, echo = FALSE, warning = FALSE}

```


# Local evaporation line model

## Define model function

*Copied to `evaporation-model.R` script for use in interactive / publication version.*

The function produces a plot-able data frame of d18O and dD values representing isotope composition of evaporate water given isotope values of input water and atmospheric conditions. Formulas and relationships are defined in @gibson2016 and @horita1994.
```{r define-model}
model.LEL <- function(env.input){
  
  # Celsius to Kelvin conversion
  TK <- env.input["TC"] + 273.15
  
  # liquid-vapor fractionation factors (Horita & Wesolowski 1994)
  aD <- exp( ((1158.8 * TK^3)/1e12) - ((1620.1 * TK^2)/1e9) + ((794.84 * TK)/1e6) - (161.04/1e3)  + (2.992e6/TK^3) )
  a18O <- exp( (-7.685e-3) + (6.7123 / TK) - (1.6664e3 / TK^2) + (0.35041e6 / TK^3) )
  
  # equilibrium isotopic separation (ibid.)
  eD <- (aD - 1) * 1000
  e18O <- (a18O - 1) * 1000
  
  # diffusion controlled fractionation (ibid.)
  ekD <- 12.5 * (1 - env.input["rh"])
  ek18O <- 14.2 * (1 - env.input["rh"])
  
  # atmospheric isotope ratios (Gibson et al 2016 eqn 18) 
  dDA <- (env.input["dDp"] - (env.input["k"] * eD)) / 
    (1 + (1e-3 * env.input["k"] * eD))
  d18OA <- (env.input["d18Op"] - (env.input["k"] * e18O)) / 
    (1 + (1e-3 * env.input["k"] * e18O))
  
  # limiting isotope ratios (Gibson et al 2016 eqn 7)
  dstarD <- ((env.input["rh"] * dDA) + ekD + (eD / aD)) / 
    (env.input["rh"] - (1e-3 * (ekD + (eD / aD))))
  dstar18O <- ((env.input["rh"] * d18OA) + ek18O + (e18O / a18O)) / 
    (env.input["rh"] - (1e-3 * (ek18O + (e18O / a18O))))
  
  # temporal enrichment slope (Gibson et al 2016 eqn 6)
  mD <- (env.input["rh"] - (1e-3 * (ekD + (eD / aD)))) /
    ((1 - env.input["rh"]) + (1e-3 * ekD))
  m18O <- (env.input["rh"] - (1e-3 * (ek18O + (e18O / a18O)))) / 
    ((1 - env.input["rh"]) + (1e-3 * ek18O))
  
  # create values for evaporation to inflow ratio (x = E / I) from 0 (no evaporation, lake growing) to 1 (fully evaporated lake)
  # adjust "by" param up or down to have fewer or more points on the model line
  x <- seq(from = 0, to = 1, by = 0.07) 
  
  # lake water isotopes for values of x (Gibson et al 2016 eqn 10) and vapor isotopes (eqn 3)
  dDL <- (env.input["dDi"]  + (mD * x * dstarD)) / (1 + mD * x)
  d18OL <- (env.input["d18Oi"] + (m18O * x * dstar18O)) / (1 + m18O * x)
  dDv <- (((dDL - eD) / aD) - (env.input["rh"] * dDA) - ekD) / (1 - env.input["rh"] + (1e-3 * ekD))
  d18Ov <- (((d18OL - e18O) / a18O) - (env.input["rh"] * d18OA) - ek18O) / (1 - env.input["rh"] + (1e-3 * ek18O))
  
  # combine modeled lake and vapor delta values in data frame for plotting
  dL <- as.data.frame(cbind("d18O" = d18OL, "dD" = dDL))
  dv <- as.data.frame(cbind("d18O" = d18Ov, "dD" = dDv))
  model.line <- as.data.frame(rbind(dL, dv))
  
  return(model.line)
}
```

## Model inputs

Build model evaporation lines by varying environmental conditions.
```{r model-inputs-single, echo = FALSE}
env.conditions <- c(
  
  # average air temperature (C); evaporation model is most accurate given an air-water interface temperature 
  "TC" = 30,
  
  # average humidity (%rh); evaporation model is most accurate given a value normalized to water surface temperature
  "rh" = 0.25,
  
  # d18O & dD (permil VSMOW) for lake water input 
  "d18Oi" = -1,
  "dDi"   = 5,
  
  # d18O & dD (permil VSMOW) for precipitation
  "d18Op" = 0,
  "dDp"   = 10,
  
  # degree to which there is isotopic equilibrium between precipitation and atmospheric water vapor
  "k" = 0.1
)

result.LEL <- model.LEL(env.conditions)
```

```{r model-inputs-vary}
# Create model scenarios for each variable
values.TC = c(25, 30, 35, 40)
values.rh = c(0.2, 0.35, 0.5, 0.65)
values.k = c(0.1, 0.3, 0.5, 0.7)

vary.TC = NULL
for (nn in 1:4){
  this.LEL = model.LEL(c(
    "TC" = values.TC[nn],
    "rh" = .25,
    "d18Oi" = -1,
    "dDi" = 5,
    "d18Op" = 0,
    "dDp" = 10,
    "k" = 0.1)) |>
    mutate(Scenario = as.character(values.TC[nn]))
  vary.TC <- rbind(vary.TC, this.LEL)
}

vary.rh = NULL
for (nn in 1:4){
  this.LEL = model.LEL(c(
    "TC" = 30,
    "rh" = values.rh[nn],
    "d18Oi" = -1,
    "dDi" = 5,
    "d18Op" = 0,
    "dDp" = 10,
    "k" = 0.1)) |>
    mutate(Scenario = as.character(values.rh[nn]))
  vary.rh <- rbind(vary.rh, this.LEL)
}

vary.k = NULL
for (nn in 1:4){
  this.LEL = model.LEL(c(
    "TC" = 30,
    "rh" = .25,
    "d18Oi" = -1,
    "dDi" = 5,
    "d18Op" = 0,
    "dDp" = 10,
    "k" = values.k[nn])) |>
    mutate(Scenario = as.character(values.k[nn]))
  vary.k <- rbind(vary.k, this.LEL)
}
```

### Model with meteoric waters
```{r plot-overlayed-scenarios, echo = FALSE, warning=FALSE}
# Multiple scenarios on one plot  

# Variable temperature ----
scenario.TC <- ggplot(data = vary.TC, aes(x = d18O, y = dD, color = Scenario)) +
  
  # Plot elements
  geom_abline(data = NULL, aes(slope = 8, intercept = 10, color = "GMWL")) +
  geom_point(data = waters.meteoric,
             aes(x = d18O, y = dD, color = WaterType, shape = WaterType), size = size.st) +
  geom_point(shape = shape.model) + geom_line() + 
  geom_label_repel(data = vary.TC |>
                   filter(d18O > 10.2),
             aes(label = Scenario)) +
  
  # Legend
  theme(legend.position = "none") +

  scale_shape_manual(
    breaks = c("Lake", "Restricted Lake", "River", "Precipitation", "Surface", "Modeled LEL", "GMWL"),
    values = c(
      "Lake" = shape.lake,
      "Restricted Lake" = shape.surface.lake,
      "River" = shape.river,
      "Precipitation" = shape.precip,
      "Surface" = shape.surface,
      "Modeled LEL" = shape.model,
      "GMWL" = 1)) +
  
  scale_color_manual(
    breaks = c("Lake", "Restricted Lake", "River", "Precipitation", "Surface", "25C", "30C", "35C", "40C", "GMWL"),
    values = c(
      "Lake" = color.lake,
      "Restricted Lake" = color.lake,
      "River" = color.river,
      "Precipitation" = color.precip,
      "Surface" = color.surface,
      "25C" = color.scenario1,
      "30C" = color.scenario2,
      "35C" = color.scenario3,
      "40C" = color.scenario4,
      "GMWL" = color.GMWL)) +
  
  labs(title = "Variable temperature",
       subtitle = "25% rh, k = 0.1", 
       x = axis.x.label, 
       y = axis.y.label) +
  
  lims(x = c(-5, 12),
       y = c(-30, 60))

plot(scenario.TC)

# Variable rh ----
scenario.rh <- ggplot(data = vary.rh, aes(x = d18O, y = dD, color = Scenario)) +
  
  # Plot elements
  geom_abline(data = NULL, aes(slope = 8, intercept = 10, color = "GMWL")) +
  geom_point(data = waters.meteoric,
             aes(x = d18O, y = dD, color = WaterType, shape = WaterType), size = size.st) +
  geom_point(shape = shape.model, show.legend = FALSE) + geom_line(show.legend = FALSE) + 
  geom_label_repel(data = vary.rh |>
                   filter(between(d18O, 9, 10.2)),
             aes(label = Scenario)) +
  
  # Legend
  theme(legend.position = "none") +

  scale_shape_manual(
    breaks = c("Lake", "Restricted Lake", "River", "Precipitation", "Surface", "Modeled LEL", "GMWL"),
    values = c(
      "Lake" = shape.lake,
      "Restricted Lake" = shape.surface.lake,
      "River" = shape.river,
      "Precipitation" = shape.precip,
      "Surface" = shape.surface,
      "Modeled LEL" = shape.model,
      "GMWL" = 1)) +
  
  scale_color_manual(
    breaks = c("Lake", "Restricted Lake", "River", "Precipitation", "Surface", "20% rh", "35% rh", "50% rh", "65% rh", "GMWL"),
    values = c(
      "Lake" = color.lake,
      "Restricted Lake" = color.lake,
      "River" = color.river,
      "Precipitation" = color.precip,
      "Surface" = color.surface,
      "20% rh" = color.scenario1,
      "35% rh" = color.scenario2,
      "50% rh" = color.scenario3,
      "65% rh" = color.scenario4,
      "GMWL" = color.GMWL)) +
  
  labs(title = "Variable humidity",
       subtitle = "30C, k = 0.1", 
       x = axis.x.label, 
       y = axis.y.label) +
  
  lims(x = c(-5, 12),
       y = c(-30, 60))

plot(scenario.rh)

# Variable k ----
scenario.k <- ggplot(data = vary.k, aes(x = d18O, y = dD, color = Scenario)) +
  
  # Plot elements
  geom_abline(data = NULL, aes(slope = 8, intercept = 10, color = "GMWL")) +
  geom_point(data = waters.meteoric,
             aes(x = d18O, y = dD, color = WaterType, shape = WaterType), size = size.st) +
  geom_point(shape = shape.model) + geom_line() + 
  geom_label_repel(data = vary.k |>
                   filter(between(d18O, 9, 10)),
             aes(label = Scenario)) +
  
  # Legend
  theme(legend.position = "none") +

  scale_shape_manual(
    breaks = c("Lake", "Restricted Lake", "River", "Precipitation", "Surface", "Modeled LEL", "GMWL"),
    values = c(
      "Lake" = shape.lake,
      "Restricted Lake" = shape.surface.lake,
      "River" = shape.river,
      "Precipitation" = shape.precip,
      "Surface" = shape.surface,
      "Modeled LEL" = shape.model,
      "GMWL" = 1)) +
  
  scale_color_manual(
    breaks = c("Lake", "Restricted Lake", "River", "Precipitation", "Surface", "0.1", "0.3", "0.5", "0.7", "GMWL"),
    values = c(
      "Lake" = color.lake,
      "Restricted Lake" = color.lake,
      "River" = color.river,
      "Precipitation" = color.precip,
      "Surface" = color.surface,
      "0.1" = color.scenario1,
      "0.3" = color.scenario2,
      "0.5" = color.scenario3,
      "0.7" = color.scenario4,
      "GMWL" = color.GMWL)) +
  
  labs(title = "Variable k",
       subtitle = "30C, 35% rh", 
       x = axis.x.label, 
       y = axis.y.label) +
  
  lims(x = c(-5, 12),
       y = c(-30, 60))

plot(scenario.k)
```

```{r plot-LEL-extra, echo = FALSE, include = FALSE}
# Single plot
plot.LEL <- ggplot(data = result.LEL, aes(x = d18O, y = dD, color = "Modeled LEL")) +
  
  # plot elements
  geom_point(shape = shape.model) + geom_line() +
  geom_point(data = waters.meteoric,
             aes(x = d18O, y = dD, color = WaterType, shape = WaterType), size = size.st, show.legend = FALSE) +
  geom_abline(data = NULL, aes(slope = 8, intercept = 10, color = "GMWL"), show.legend = FALSE) +
  
  # legend
  scale_shape_manual(
    breaks = c("Lake", "Restricted Lake", "River", "Precipitation", "Surface", "Modeled LEL", "GMWL"), 
    # set plot element shapes
    values = c(
      "Lake" = shape.lake,
      "Restricted Lake" = shape.surface.lake,
      "River" = shape.river,
      "Precipitation" = shape.precip,
      "Surface" = shape.surface,
      "Modeled LEL" = shape.model,
      "GMWL" = 1)) +
  
  scale_color_manual(
    breaks = c("Lake", "Restricted Lake", "River", "Precipitation", "Surface", "Modeled LEL", "GMWL"), 
    # set plot element colors
    values = c(
      "Lake" = color.lake,
      "Restricted Lake" = color.lake,
      "River" = color.river,
      "Precipitation" = color.precip,
      "Surface" = color.surface,
      "Modeled LEL" = color.model,
      "GMWL" = color.GMWL)) +
  
  guides(color = guide_legend(override.aes=list(
    linetype = c(rep("blank", 5), rep("solid", 2)),
    shape = c(shape.lake, shape.surface.lake, shape.river, shape.precip, shape.surface, shape.model, NA),
    size = 2))) +
  
  labs(x = axis.x.label, 
       y = axis.y.label,
       caption = paste(names(env.conditions), "=", env.conditions, collapse = ", "))

plot(plot.LEL)

# PScenario panels  
panel.LEL <- ggplot(data = vary.TC, aes(x = d18O, y = dD)) +
  
  # plot elements
  geom_point(shape = shape.model) + geom_line() +
  geom_point(data = waters.meteoric,
             aes(x = d18O, y = dD, color = WaterType, shape = WaterType), size = size.st, show.legend = FALSE) +
  geom_abline(data = NULL, aes(slope = 8, intercept = 10, color = "GMWL"), show.legend = FALSE) +
  
  # legend
  scale_shape_manual(
    breaks = c("Lake", "Restricted Lake", "River", "Precipitation", "Surface", "Modeled LEL", "GMWL"), 
    # set plot element shapes
    values = c(
      "Lake" = shape.lake,
      "Restricted Lake" = shape.surface.lake,
      "River" = shape.river,
      "Precipitation" = shape.precip,
      "Surface" = shape.surface,
      "Modeled LEL" = shape.model,
      "GMWL" = 1)) +
  
  # scale_color_manual(
  #   breaks = c("Lake", "Restricted Lake", "River", "Precipitation", "Surface", "Modeled LEL", "GMWL"), 
  #   # set plot element colors
  #   values = c(
  #     "Lake" = color.lake,
  #     "Restricted Lake" = color.lake,
  #     "River" = color.river,
  #     "Precipitation" = color.precip,
  #     "Surface" = color.surface,
  #     "Modeled LEL" = color.model,
  #     "GMWL" = color.GMWL)) +
  # 
  # guides(color = guide_legend(override.aes=list(
  #   linetype = c(rep("blank", 5), rep("solid", 2)),
  #   shape = c(shape.lake, shape.surface.lake, shape.river, shape.precip, shape.surface, shape.model, NA),
  #   size = 2))) +
  
  labs(x = axis.x.label, 
       y = axis.y.label) +

  facet_grid(.~Scenario)          

plot(panel.LEL)
```

### Test models

Set k = 0.1, equilibrium between precip and vapor above the lake is probably low because vapor isotope composition reflects evaporated lake water, not precip signal, Changing k has little effect on model slope. 

Set TC = 30, close to mean of 11 measured surface temperatures and close to recorded MAAT. Temperature also has little effect on model slope. 

- Is the best model the one that best fits the lake data? 
- Regress lake only
- Compare models with lake regression
- Quantify effect of each atmospheric variable on model — how? 



## Mass balance

Estimate lake volumes from [USDA Foreign Agriculture Service](https://ipad.fas.usda.gov/cropexplorer/global_reservoir/gr_regional_chart.aspx?regionid=eafrica&reservoir_name=Turkana&lakeid=000093) and estimate percent increases in inputs needed.