---
title: "Lake water evaporation model and mass balance calculations for Lake Turkana, Kenya"
date: 2021-11-03
output: 
  html_document: 
    highlight: monochrome
  html_notebook: 
    highlight: monochrome
editor_options: 
  chunk_output_type: inline
reference-section-title: "References"
---

```{r, setup, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(here, knitr, tidyverse, kableExtra, ggthemes)
```

# Local evaporation line model

## Define model function

The function produces a plot-able data frame of δ^18^O and δD values representing isotope composition of evaporate water given isotope values of input water and atmospheric conditions. Formulas and relationships are defined in @gibson2016 and @horita1994.

```{r define-model}
model.LEL <- function(env.input){
  
  # Celsius to Kelvin conversion
  TK <- env.input["TC"] + 273.15
  
  # liquid-vapor fractionation factors (Horita & Wesolowski 1994)
  aD <- exp( ((1158.8 * TK^3)/1e12) - ((1620.1 * TK^2)/1e9) + ((794.84 * TK)/1e6) - (161.04/1e3)  + (2.992e6/TK^3) )
  a18O <- exp( (-7.685e-3) + (6.7123 / TK) - (1.6664e3 / TK^2) + (0.35041e6 / TK^3) )
  
  # equilibrium isotopic separation (ibid.)
  eD <- (aD - 1) * 1000
  e18O <- (a18O - 1) * 1000
  
  # diffusion controlled fractionation (ibid.)
  ekD <- 12.5 * (1 - env.input["rh"])
  ek18O <- 14.2 * (1 - env.input["rh"])
  
  # atmospheric isotope ratios (Gibson et al 2016 eqn 18) 
  dDA <- (env.input["dDp"] - (env.input["k"] * eD)) / 
    (1 + (1e-3 * env.input["k"] * eD))
  d18OA <- (env.input["d18Op"] - (env.input["k"] * e18O)) / 
    (1 + (1e-3 * env.input["k"] * e18O))
  
  # limiting isotope ratios (Gibson et al 2016 eqn 7)
  dstarD <- ((env.input["rh"] * dDA) + ekD + (eD / aD)) / 
    (env.input["rh"] - (1e-3 * (ekD + (eD / aD))))
  dstar18O <- ((env.input["rh"] * d18OA) + ek18O + (e18O / a18O)) / 
    (env.input["rh"] - (1e-3 * (ek18O + (e18O / a18O))))
  
  # temporal enrichment slope (Gibson et al 2016 eqn 6)
  mD <- (env.input["rh"] - (1e-3 * (ekD + (eD / aD)))) /
    ((1 - env.input["rh"]) + (1e-3 * ekD))
  m18O <- (env.input["rh"] - (1e-3 * (ek18O + (e18O / a18O)))) / 
    ((1 - env.input["rh"]) + (1e-3 * ek18O))
  
  # create values for evaporation to inflow ratio (x = E / I) from 0 (no evaporation, lake growing) to 1 (fully evaporated lake)
  # adjust "by" param up or down to have fewer or more points on the model line
  x <- seq(from = 0, to = 1, by = 0.07) 
  
  # lake water isotopes for values of x (Gibson et al 2016 eqn 10) and vapor isotopes (eqn 3)
  dDL <- (env.input["dDi"]  + (mD * x * dstarD)) / (1 + mD * x)
  d18OL <- (env.input["d18Oi"] + (m18O * x * dstar18O)) / (1 + m18O * x)
  dDv <- (((dDL - eD) / aD) - (env.input["rh"] * dDA) - ekD) / (1 - env.input["rh"] + (1e-3 * ekD))
  d18Ov <- (((d18OL - e18O) / a18O) - (env.input["rh"] * d18OA) - ek18O) / (1 - env.input["rh"] + (1e-3 * ek18O))
  
  # combine modeled lake and vapor delta values in data frame for plotting
  dL <- as.data.frame(cbind("d18O" = d18OL, "dD" = dDL))
  dv <- as.data.frame(cbind("d18O" = d18Ov, "dD" = dDv))
  model.line <- as.data.frame(rbind(dL, dv))
  
  return(model.line)
}
```

## Model inputs

Build model evaporation lines using environmental conditions The resulting evaporation line represents these inputs given to the function above.

```{r model-inputs-1}
env.conditions <- c(
  
  # average air temperature (C); evaporation model is most accurate given an air-water interface temperature 
  "TC" = 28,
  
  # average humidity (%rh); evaporation model is most accurate given a value normalized to water surface temperature
  "rh" = 0.25,
  
  # d18O & dD (permil VSMOW) for lake water input 
  "d18Oi" = -0.4,
  "dDi"   = 9,
  
  # d18O & dD (permil VSMOW) for precipitation
  "d18Op" = 0.9,
  "dDp"   = 10,
  
  # degree to which there is isotopic equilibrium between precipitation and atmospheric water vapor
  "k" = 0.1
)

result.LEL <- model.LEL(env.conditions)
```

## Plot model with data

### Load and filter data (this study)

*To do:* Talk about the "Evaporated Surface" data points --- all but one are from TBI-Ileret and appear on a line in plot 1; the last is from a pond in Topernawi laga. We should try to find out more about the Ileret conditions and what they capture; the line looks roughly parallel to the LEL. What is the relevance of the Topernawi data point? Is it worth comparing one point each from Topernawi laga and lake shore?

```{r waters-all, echo = FALSE}
waters.all <- read.csv(here("data/2016-2020_waters.csv"))

waters.meteoric <- waters.all |> 
  filter(WaterType == "Lake" | WaterType == "River" | WaterType == "Precipitation")

kable(waters.all |> 
      select(SampleID, Date, WaterType, Location, d18O, dD, Dexcess),
      format="html") |> 
      kable_styling() |> 
      scroll_box(height="220px", width="100%")
```

### Define plot aesthetics

Colorblind-friendly palette from @tol2018.

```{r plot-aesthetics, echo = FALSE}
theme_set(theme_tufte(base_family = "Helvetica"))

axis.x.max <- max(waters.all$d18O) + 0.5
axis.x.min <- min(waters.all$d18O) - 0.5
axis.y.max <- max(waters.all$dD) + 0.5
axis.y.min <- min(waters.all$dD) - 0.5

axis.x.label <- expression(paste(delta^{18}, "O (‰ VSMOW)"))
axis.y.label <- expression(paste(delta, "D (‰ VSMOW)"))

color.lake <- "#228833"
color.precip <- "#66CCEE"
color.river <- "#4477AA"
color.ground <- "#EE6677"
color.surface <- "#CCBB44"
color.GMWL <- "gray19"
color.model <- "gray69"

shape.lake <- 16 # circle
shape.precip <- 17 # triangle
shape.river <- 15 # square
shape.ground.deep <- 7 # x in box
shape.ground.shallow <- 4 # x
shape.surface.evap <- 5 # diamond empty
shape.surface.lake <- 1 # circle empty
shape.model <- 3 # cross 
```

### Model with meteoric waters

Local evaporation model for environmental conditions: `r paste(names(env.conditions), "=", env.conditions, collapse=", ")`

```{r LEL-meteoric, echo = TRUE}
plot.LEL <- ggplot(data = result.LEL, aes(x = d18O, y = dD, color = "Modeled LEL")) +
  
  # plot elements
  geom_point(shape = shape.model) + geom_line() +
  geom_point(data = waters.meteoric, aes(x = d18O, y = dD, color = WaterType, shape = WaterType), size = 2, show.legend = FALSE) +
  geom_abline(data = NULL, aes(slope = 8, intercept = 10, color = "GMWL"), show.legend = FALSE) +
  
  # legend
  scale_shape_manual(
    breaks = c("Lake", "River", "Precipitation", "Modeled LEL", "GMWL"), 
    # set plot element shapes
    values = c(
      "Lake" = shape.lake,
      "River" = shape.river,
      "Precipitation" = shape.precip,
      "Modeled LEL" = shape.model,
      "GMWL" = 1)) +
  
  scale_color_manual(
    breaks = c("Lake", "River", "Precipitation", "Modeled LEL", "GMWL"), 
    # set plot element colors
    values = c(
      "Lake" = color.lake,
      "River" = color.river,
      "Precipitation" = color.precip,
      "Modeled LEL" = color.model,
      "GMWL" = color.GMWL)) +
  
  guides(color = guide_legend(override.aes=list(
    linetype = c(rep("blank", 3), rep("solid", 2)),
    shape = c(shape.lake, shape.river, shape.precip, shape.model, NA),
    size = 2))) +
  
  # axes and titles
  coord_cartesian(xlim = c(-5,9), ylim = c(-25,50)) + 
  theme(legend.title = element_blank(), 
        panel.border = element_rect(fill = NA)) + 
  labs(x = axis.x.label, 
       y = axis.y.label,
       caption = paste(names(env.conditions), "=", env.conditions, collapse = ", "))

plot(plot.LEL)
```
*To do:* Generate a 4 or 6 panel figure with four sets of input values. Use `apply()` for this? 


## Test models

*To do:* Get statistical comparison (ANOVA? T-test?) between each model LEL and the lake waters

### De-trend lake data

*To do:* Subtract LEL model from lake data to check for fit

Like this?

-   Get linear equation for LEL

-   Apply linear function to lake δ^18^O values to produce data frame of real δ^18^O vs modeled δD

-   Use `detrend()` to remove linear trend from new data frame

## All data

*To do:* Add @levin2009

```{r plot-all, echo = FALSE}
plot.all <- ggplot(data = waters.all, aes(x = d18O, y = dD, shape = WaterType, color = WaterType), show.legend = FALSE) +
  geom_point() +
  geom_abline(data = NULL, aes(slope = 8, intercept = 10, color = "GMWL")) +

  # legend
  scale_shape_manual(
    breaks = c("Lake", "River", "Precipitation", "Shallow Ground", "Deep Ground", "Evaporated Surface", "Restricted Lake", "GMWL"), 
    # set plot element shapes
    values = c(
      "Lake" = shape.lake,
      "River" = shape.river,
      "Precipitation" = shape.precip,
      "Shallow Ground" = shape.ground.shallow,
      "Deep Ground" = shape.ground.deep,
      "Evaporated Surface" = shape.surface.evap,
      "Restricted Lake" = shape.surface.lake,
      "GMWL" = 1)) +
  
  scale_color_manual(
    breaks = c("Lake", "River", "Precipitation", "Shallow Ground", "Deep Ground", "Evaporated Surface", "Restricted Lake", "GMWL"),
    # set plot element colors
    values = c(
      "Lake" = color.lake,
      "River" = color.river,
      "Precipitation" = color.precip,
      "Shallow Ground" = color.ground,
      "Deep Ground" = color.ground,
      "Evaporated Surface" = color.surface,
      "Restricted Lake" = color.lake,
      "GMWL" = color.GMWL)) +
  
  guides(shape = "none", color = guide_legend(override.aes = list(
    linetype = c(rep("blank", 7), "solid"),
    shape = c(shape.lake, shape.river, shape.precip, shape.ground.shallow, shape.ground.deep, shape.surface.evap, shape.surface.lake, NA),
    size = 3))) +
  
  # axes and titles
  coord_cartesian(xlim = c(axis.x.min, axis.x.max), ylim = c(axis.y.min, axis.y.max)) + 
  theme(legend.title = element_blank(), 
        panel.border = element_rect(fill = NA)) + 
  labs(x = axis.x.label, 
       y = axis.y.label)

plot(plot.all)
```

ggplot mystery: Why is the legend symbol for GMWL like that?

## Mass balance

Estimate lake volumes from [USDA Foreign Agriculture Service](https://ipad.fas.usda.gov/cropexplorer/global_reservoir/gr_regional_chart.aspx?regionid=eafrica&reservoir_name=Turkana&lakeid=000093) and estimate percent increases in inputs needed.